Data Exploration, Quality Control, and Statistical Analysis of ChIP-exo/nexus Experiments

### 2016-07-05 paper/chipexo_QC_vs2.tex
previous version of the paper

\usepackage{xcolor}
\usepackage{url}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{tikz}
\usetikzlibrary{shapes,arrows}
\usepackage{float}
\usepackage{verbatim}
\usepackage{tabu}
\usepackage{multirow}
\usepackage{framed}

@@

\newcommand{\pname}[1]{\texttt{ChIPexoQual}}
\newcommand{\SK}[1]{\textcolor{red}{SK: #1}}

@@

    We evaluated our QC pipeline with both  large collections of public ChIP-exo/nexus data and 
    multiple, new ChIP-exo datasets from \textit{E. Coli}.  \pname{} analysis of these datasets resulted in  guidelines  for
   using these QC metrics across a wide range of sequencing depths 
   and further insights for modeling ChIP-exo data. %In addition, we demonstrate that the
    %ChIP-exo QC pipeline it is also applicable to ChIP-nexus data,
    %showing that those experiments present higher quality than
    %ChIP-exo experiments under similar conditions.
    Finally, although ChIP-exo experiments have been compared to ChIP-seq experiments with single-end (SE) sequencing, we provide, for the first time, comparisons with paired-end (PE)  ChIP-seq experiments.
    %We compared ChIP-exo with Paired End (PE) and Single End (SE)
    %ChIP-Seq and found the following characteristics: First, although
    %often assumed in ChIP-exo data analysis methods, the ``peak pair''
    %assumptions does not hold locally in actual ChIP-exo data. Second,
    %we for the first time compared PE ChIP-Seq with ChIP-exo and found
%    that 
    We illustrate that, at fixed sequencing depths, ChIP-exo provides higher
    sensitivity, specificity, and spatial resolution than PE
    ChIP-seq and both significantly outperform their SE ChIP-seq counterpart. Furthermore, we show that ChIP-exo and PE ChIP-seq are
    comparable in sensitivity for closely located binding events, but
    as the average distance between binding events increases, ChIP-exo
    exhibits higher sensitivity than PE ChIP-Seq. \SK{The last sentence does not really make much sense? }

@@

Chromatin Immunoprecipitation followed by exonuclease
digestion and next generation sequencing (ChIP-exo) 
is currently one of the state-of-the-art high throughput assays for profiling protein-DNA interactions at or close to
single base-pair resolution \cite{exo1}. It presents a powerful alternative to popularly used ChIP-seq
(Chromatin immunoprecipitation coupled with next generation
sequencing) assay. ChIP-exo experiments first capture millions of DNA
fragments ($150$ - $250$ bps in length) that the protein under study
interacts with using  a
protein-specific antibody and random fragmentation of DNA. Then, $\lambda$-exonuclease ($\lambda$-exo) is deployed to trim the
$5^{\prime}$ end of each DNA fragment to each protein-DNA interaction boundary. This step is unique to ChIP-exo and
aims to achieve  significantly higher spatial resolution
compared to ChIP-seq. Finally, high throughput sequencing of a small
region ($36$ to $100$ bps) at the $5^{\prime}$ end of each fragment generates
millions of reads. 
 Similarly, ChIP-nexus (Chromatin Immunoprecipitation followed by exonuclease
digestion, unique barcode, single ligation and next generation
ligation)  \cite{chipnexus} is a further modification on the
ChIP-exo protocol. ChIP-nexus aims to overcome limitations of ChIP-exo by 
by yielding high complexity libraries with numbers of cells comparable to that of ChIP-seq experiments. This is achieved by reducing the numbers of ligations in the standard ChIP-exo protocol from two to one, and adding unique, randomized barcodes to adoptors to enable monitoring of overamplification. 
%where both sequencing adaptors are ligated at the
%end of the ChIP fragments. Then, after exonuclease digestion, DNA self-circularization with circLigase, and restriction enzyme cutting between the two adaptors, the final library is amplified. Compared to ChIP-exo, ChIP-nexus aims to   attain  higher resolution analysis while yielding higher complexity libraries.
Figure~\ref{fig:chip_diagram} illustrates the
differences between different ChIP-based protocols: ChIP-exo, single-end (SE) ChIP-seq, paired-end
(PE) ChIP-seq, ChIP-nexus. The $5^{\prime}$ ends of a ChIP-exo/nexus experiment are clustered
more tightly around the binding sites of the protein than in a ChIP-seq
experiment. In a PE ChIP-seq experiment, both ends are sequenced as opposed to only the $5^{\prime}$ end
in a SE ChIP-seq. \SK{Note to Rene: ChIP-Nexus paper has a good description of what can go wrong with ChIP-exo; It seems like most of the read imbalance could also be due to ligation inefficiency. Although this probably does not explain why we have reads only from one strand in some regions. This is more likely explained by over digestion in one strand or single stranded TF-DNA interaction. Think a bit more on these issues and understand how the ligation could give rise to strand imbalance, we can ask Bob to have a closer look at the section where we have this discussion.}

@@

% while there are not established QC pipelines for ChIP-exo; previous ChIP-exo analyses used ChIP-Seq samples to compare the resolution between experiments \cite{exo1,exoillumina,exo2}.

@@

\subsection*{Publicly available ChIP-exo/nexus and novel \textit{E. coli} ChIP-seq/exo datasets}
We utilized a rich collection of publicly available ChIP-exo/nexus data from multiple organisms to build and evaluate our quality control pipeline (Table~\ref{tab:qc}). These include:   CTCF factor
in human HeLa cells \cite{exo1}; ER factor in human MCF-7
cells \cite{exoillumina}; GR factor in IMR90, K562, and U2OS human
cells \cite{starick15}; TBP factor in human K562 cells
\cite{venters13}. 
ChIP-nexus data included experiments from \cite{chipnexus} profiling
TBP in human K562 cells, MyC and Max in \emph{D. Melanogaster} S2 cells, and Twist and
Dorsal in \emph{D. Melanogaster} embryo.

In order to have a setting where we can compare SE and PE ChIP-seq with their ChIP-exo counterpart, we profiled 
$\sig$ under a variety of  conditions in \textit{E. coli} with both ChIP-exo (Table~\ref{tab:qc_sig}) and PE ChIP-seq (\SK{Table SXX}). Collectively, we 
 generated $\sig$ factor ChIP-exo, PE and SE ChIP-seq experiments  under aerobic ($+O_2$) and anaerobic ($-O_2$)
conditions in glucose minimal media. Similarly, we generated $\sig$ factor ChIP-exo, PE and
SE (generated \emph{in silico}) ChIP-seq experiments in \emph{E. Coli}
under aerobic ($+O_2$) conditions with  and without  rifampicin  treatment.
We refer to these latter set of experiments  as time point 20 minutes and time point 0 minutes \SK{Check for consistency with other parts of the paper and correct, ok to use different naming convention}.
We specifically used these experimental data  for comparisons of ChIP-exo and PE
ChIP-seq assays in identifying closely spaced binding events and their resolution, i.e., physical proximity of the predicted events to the actual binding sites.  This comparison benefits from using $\sig$  which is a transcription initiation factor of housekeeping genes
in \emph{E. Coli}. Many \textit{E. coli} promoters contain
multiple transcription start sites (TSS). These TSSs are often
closely spaced, i.e., within 10 $\sim$ 150 bps of each other, and are considered  to be multiple ``switches'' that differentially
regulate gene expression under diverse growth conditions
\cite{regulondb}.

@@

\textit{Read distributions within signal and background regions.}
We first compared ChIP-seq and ChIP-exo in terms of data features that are well studied in ChIP-seq studies.
Our $\sig$  ChIP-seq and ChIP-exo samples from \textit{E. coli} are especially well suited for this task since they are all deeply sequenced compared to the genome size of \textit{E. coli}.   Figure~\ref{fig:comp} summarizes this comparison for one biological replicate  of ChIP-exo and ChIP-seq experiments from the same biological conditions (samples x and y from Table~\ref{tab:qc_sig} \SK{Maybe add a column as sample number to this table to make referencing easy}). Comparisons with other paired \textit{E. coli} ChIP-seq and ChIP-exo  samples led to similar conclusions (Supplementary Figure \SK{SXXX}).
%compared various factors that could affect binding site
%identification between ChIP-exo and ChIP-Seq data by using the
%ChIP-exo reads from the first biological sample and first replicate
%grown under aerobic condition (first line from Table \ref{tab:qc_sig})
%and a SE ChIP-Seq replicate grown under the same conditions. 
We summarized the extended raw read counts within $150$ bps non-overlapping intervals, i.e., bins, interrogating the genome.
%In order
%to compare distribution of signal and background between ChIP-exo and
%ChIP-Seq data, we counted the number of extended reads mapping to a
%partition of the genome into non-overlapping bins. 
Figure~\ref{fig:comp}(A) depicts that, as expected,  ChIP read counts from ChIP-exo and ChIP-seq are linearly 
correlated especially at high read counts. This indicates that signals for potential binding
sites are well reproducible between ChIP-exo and ChIP-seq data. 
In contrast, there is a clear difference
among the two data types for bins with low read counts, highlighting expected differences in the  background read distributions 
of the two data types. 
\SK{***After seeing these types of plots from other samples (e..g., ER in MCF7, it is unlikely that what we are seeing from $\sig$ samples regarding background reads (the next few sentences) is typical. For CTCF and $\sig$, the ChIP-exo samples have a wider dynamic range, but this could be due to differences in sequencing depth. *** Specifically, background reads from ChIP-seq are  almost
uniformly distributed over background (non-enriched) regions. In comparison, background regions in ChIP-exo show larger variation but have overall lower read counts. Furthermore, there are considerably large numbers of bins with zero read counts in ChIP-exo and non-zero read counts in ChIP-seq.
%and the
%ChIP tag counts in there regions were significantly larger than
%zero. In contrast, in ChIP-exo data, there was larger variation in
%ChIP tag counts among background regions and ChIP tag counts were much
%lower in these regions compared to ChIP-Seq data. There were also
%large proportion of regions without any read in ChIP-exo data. 
Overall, these observations indicate that  a much smaller portion of the
genome is expected to be background in ChIP-exo compared to ChIP-seq and 
 methods that
specifically model the background read distribution might benefit from acknowledging this.}

\textit{Peak-pair assumption.} We next evaluated the peak-pair
assumption, i.e., a cluster of reads
in the forward strand is usually paired with a cluster of reads in the
reverse strand that is located on the right-hand-side of  the binding
site, that is commonly utilized in designing statistical analysis methods for ChIP-exo data \cite{mace, cexor, peakzilla}.
We considered the set of peaks identified in both the ChIP-seq and ChIP-exo samples as high quality peaks (Materials and Methods) and calculated the proportion of forward strand reads in these regions (Figure~\ref{fig:comp}(B)). This plot reveals a higher level of strand imbalance for ChIP-exo compared to ChIP-seq. 
%In order to evaluate this assumption, we reviewed the proportion of reads in the forward strand in high quality ChIP-exo peaks such as at least one binding site is predicted in both ChIP-exo and ChIP-Seq data.
%We
%found that strands of reads were much less balanced in ChIP-exo data
%than in ChIP-Seq data in these regions with potential binding sites
%(Fig. \ref{fig:comp}B) and this indicates that the peak pair
%assumption might not hold in ChIP-exo data. 
Potential reasons for this observation include ligation efficiency, efficiency of $\lambda$-exo digestion, 
and single-stranded protein-DNA interactions. Overall, such an imbalance is prevalent in  \SK{X}\% of the ChIP-exo samples used in this paper.
%the enzyme digestion or the strength at which the protein binds to the
%DNA: Although it is expected for the exonuclease enzyme to digest the
%ChIP fragments starting by their $5\prime$ ends and stopping when
%finding a binding event, it may not be able to reach and digest every
%fragment in the ChIP sample. On the other hand, if the protein is not
%bound to both DNA strand, the enzyme may completely digest the
%fragment. Finally both effect are increased by the PCR amplification.

\textit{Mappability and GC-content bias.} We next evaluated ChIP-exo data of CTCF in HeLa cells
\cite{exo1} to investigate biases inherent to next generation sequencing experiments with eukaryotic genomes. 
Figures \ref{fig:comp}(C) and
\ref{fig:comp}(D) display the bin-level average read counts against
mappability and GC-content. Each data point is obtained by averaging
the read counts across bins with the same mappability of GC-content. 
These biases, increasing linear trend with mappability and non-linear trend with GC-content, are similar to those observed in ChIP-seq datasets \cite{benjamini2011, mosaics,quest}.
%In Figure \ref{fig:comp}C it is shown that the ChIP-exo tag
%counts linearly increases with the mappability score and in Figure
%\ref{fig:comp}D it is shown that for GC - content below 0.6, the mean
%ChIP tag count increases and for GC - content greater than 0.6 it
%shows a decreasing trend. Benjamini and Speed,
%2011\nocite{benjamini2011} and Kuan et al., 2011\nocite{mosaics}
%studied the presence of the mappability and GC - content biases in
%ChIP-Seq's background. It is not surprising to see these biases also
%present in ChIP-exo data, since ChIP-Seq and ChIP-exo signal seems to
%be linearly correlated for enriched regions (Figure
%\ref{fig:comp}A). Rozowsky et al., 2009\nocite{peakseq} and Valouev et
%al., 2008\nocite{quest} provide in depth analysis of the mappability
%and GC - content biases for ChIP-Seq respectively. 
This observation  indicates that analysis of  ChIP-exo data
should benefit from methods that take into account  apparent
sequencing biases such as mappability and GC content, mostly when an
input  control sample is not available.

@@

The Strand Cross-Correlation (SCC), introduced by \cite{strandcc}, is the most commonly used quality metric in assessing 
ChIP-seq enrichment quality. It aims to quantify  how well the reads mapped to each
strand are clustered around the locations of the protein-DNA interaction sites by calculating the Pearson correlation among forward and backward strands reads by shifting them across a range that covers both the read length of the experiment and the expected average fragment length.
Typical SCC profiles exhibit two local maxima: at the average fragment length and the read length. In high quality experiments with clear ChIP enrichment, the average fragment length maximum coincides with the global maximum.
%, and usually it is expected to observed two local
%maxima, one when the profiles are shifted by the average read length
%and another when the profiles are shifted by the unobserved fragment
%length. 
%In a high quality ChIP-Seq dataset the last one is also the
%SCC global maxima. 
In an
idealized ChIP-exo experiment where the DNA fragments are digested
 to  the boundaries of the protein-DNA interaction sites,  we would expect the SCC profile to maximize at the motif length indicating clustering of the forward and reverse strand reads around the  binding site.
 Figure~\ref{fig:scc_exo} displays the SCC curves for the CTCF  HeLa samples where the ChIP-exo curve shows  local maxima at the motif and read
lengths, while the SE ChIP-seq curves have a local maxima at the read
length and a global maxima at the average fragment
length. SCC profiles for other samples  are available in Supplementary Figures \SK{XXX}. The read length and motif length maxima are often in close proximity of each other; as a result, this renders
QC metrics such as the Normalized
Strand Cross-Correlation (NSC) or the Relative Strand
Cross-Correlation (RSC) harder to interpret; however, the profile itself seems informative about the enrichment signal in ChIP-exo/nexus experiments \SK{this seems true at least based on my initial look at the plots, but we should be cautious in reporting this}.

@@

We specifically considered recently developed Peakzilla \cite{peakzilla}, MACE
\cite{mace}, GEM \cite{gem}, and dPeak \cite{dpeak} in our evaluations with 
high quality \emph{E.Coli} samples (samples \SK{x and y} from the aerobic condition). Among these methods, Peakzilla and MACE are specifically developed for ChIP-exo analysis, whereas both dPeak and GEM can identify high resolution binding events from ChIP experiments, rendering them suitable for ChIP-exo analysis as well. 
%After we determined which biases could affect a ChIP-exo experiment's
%quality we decided to proceed and examine binding sites in high
%resolution on high quality ChIP-exo experiments. We examined first
%both aerobic replicates from the \emph{E.Coli} first biosample.

Figures~\ref{fig:methods_comp}A and \ref{fig:methods_comp}B compare
binding events identified with all the four methods  in terms of resolution,
where the resolution is defined as the distance from a RegulonDB \cite{regulondb}
annotation to its closest prediction. The resolutions of  all the methods
are comparable for the first replicate (Figure~\ref{fig:methods_comp}(A)) and dPeak, on average, has slightly better resolution than the rest in the second replicate (Figure~\ref{fig:methods_comp}(B)).

>>>>

\usepackage{xcolor}
\usepackage{url}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{tikz}
\usetikzlibrary{shapes,arrows}
\usepackage{float}
\usepackage{verbatim}
\usepackage{tabu}
\usepackage{multirow}
\usepackage{framed}
\usepackage{xr}
\externaldocument{supplement}

\usepackage{verbatim}
\usepackage{gensymb}
\usepackage{listings}

@@

\newcommand{\pname}[1]{\texttt{ChIPexoQual}}
\newcommand{\SK}[1]{\textcolor{red}{SK: #1}}
\newcommand{\RW}[1]{\textcolor{blue}{RW: #1}}

@@

    We evaluated our QC pipeline with both large collections of public
    ChIP-exo/nexus data and multiple, new ChIP-exo datasets from
    \textit{E. Coli}.  \pname{} analysis of these datasets resulted in
    guidelines for using these QC metrics across a wide range of
    sequencing depths and further insights for modeling ChIP-exo
    data. %In addition, we demonstrate that the
    %ChIP-exo QC pipeline it is also applicable to ChIP-nexus data,
    %showing that those experiments present higher quality than
    %ChIP-exo experiments under similar conditions.
    Finally, although ChIP-exo experiments have been compared to
    ChIP-seq experiments with single-end (SE) sequencing, we provide,
    for the first time, comparisons with paired-end (PE) ChIP-seq
    experiments.
    %We compared ChIP-exo with Paired End (PE) and Single End (SE)
    %ChIP-Seq and found the following characteristics: First, although
    %often assumed in ChIP-exo data analysis methods, the ``peak pair''
    %assumptions does not hold locally in actual ChIP-exo data. Second,
    %we for the first time compared PE ChIP-Seq with ChIP-exo and found
%    that 
    We illustrate that, at fixed sequencing depths, ChIP-exo provides
    higher sensitivity, specificity, and spatial resolution than PE
    ChIP-seq and both significantly outperform their SE ChIP-seq
    counterpart. Furthermore, we show that ChIP-exo and PE ChIP-seq
    are comparable in sensitivity for closely located binding events,
    but as the average distance between binding events increases,
    ChIP-exo exhibits higher sensitivity than PE ChIP-Seq.

    \SK{The last sentence does not really make much sense? }, \RW{I
      meant that when we have a peak composed by more than one binding
      event, as the mean distance between binding events increases
      then ChIP-exo is more likely to identify all the binding events,
      I used sensitivity because we defined it as the proportion of
      identified binding events in a given peak. I think it depends if
      we decide to add Fig8 in the main article or in the
      supplement.}

@@

Chromatin Immunoprecipitation followed by exonuclease digestion and
next generation sequencing (ChIP-exo) is currently one of the
state-of-the-art high throughput assays for profiling protein-DNA
interactions at or close to single base-pair resolution
\cite{exo1}. It presents a powerful alternative to popularly used
ChIP-seq (Chromatin immunoprecipitation coupled with next generation
sequencing) assay. ChIP-exo experiments first capture millions of DNA
fragments ($150$ - $250$ bps in length) that the protein under study
interacts with using a protein-specific antibody and random
fragmentation of DNA. Then, $\lambda$-exonuclease ($\lambda$-exo) is
deployed to trim the $5^{\prime}$ end of each DNA fragment to each
protein-DNA interaction boundary. This step is unique to ChIP-exo and
aims to achieve significantly higher spatial resolution compared to
ChIP-seq. Finally, high throughput sequencing of a small region ($36$
to $100$ bps) at the $5^{\prime}$ end of each fragment generates
millions of reads.  Similarly, ChIP-nexus (Chromatin
Immunoprecipitation followed by exonuclease digestion, unique barcode,
single ligation and next generation ligation) \cite{chipnexus} is a
further modification on the ChIP-exo protocol. ChIP-nexus aims to
overcome limitations of ChIP-exo by yielding high complexity libraries
with numbers of cells comparable to that of ChIP-seq experiments. This
is achieved by reducing the numbers of ligations in the standard
ChIP-exo protocol from two to one, and adding unique, randomized
barcodes to adaptors to enable monitoring of
overamplification. Figure~\ref{fig:1}A illustrates the differences between
different ChIP-based protocols: ChIP-exo, single-end (SE) ChIP-seq,
paired-end (PE) ChIP-seq, ChIP-nexus. The $5^{\prime}$ ends of a
ChIP-exo/nexus experiment are clustered more tightly around the
binding sites of the protein than in a ChIP-seq experiment. In a PE
ChIP-seq experiment, both ends are sequenced as opposed to only the
$5^{\prime}$ end in a SE ChIP-seq.

% \RW{This modification makes a library generated
%   using the ChIP-nexus protocol to be larger than one generated with
%   ChIP-exo, which }
%where both sequencing adaptors are ligated at the
%end of the ChIP fragments. Then, after exonuclease digestion, DNA self-circularization with circLigase, and restriction enzyme cutting between the two adaptors, the final library is amplified. Compared to ChIP-exo, ChIP-nexus aims to   attain  higher resolution analysis while yielding higher complexity libraries.

\SK{Note to Rene: ChIP-nexus paper has a good description of what can
  go wrong with ChIP-exo; It seems like most of the read imbalance
  could also be due to ligation inefficiency. Although this probably
  does not explain why we have reads only from one strand in some
  regions. This is more likely explained by over digestion in one
  strand or single stranded TF-DNA interaction. Think a bit more on
  these issues and understand how the ligation could give rise to
  strand imbalance, we can ask Bob to have a closer look at the
  section where we have this discussion.}

% \RW{1 - I think it may come from the starting material in each side. I
%   think that at least the number of fragments on each side must
%   preserve that if there is more starting reads in one side, then this
%   must be the case after the amplification.}

% \RW{2 - There is not control about how the enzyme digests the
%   fragments, as far as we know it starts on the $5\prime$ side and
%   stops until it finds the TF or it runs out of energy to keep
%   digesting. I think Bob mentioned that it may be possible for the
%   enzyme to pass though the TF if it is left digesting for a longer
%   time, but if that's the case then the question may be why it
%   preffers one strand over the other, which may lead to the GC-bias
%   but as far as I can tell is very hard to observe that because that
%   bias appear also in ChIP-seq (and it seeems that also appear in the
%   circularization step from ChIP-nexus).}

% \RW{3 - I think that the regions formed by fragments in one strand,
%   may be copies of few undigested fragments, in a good experiment the
%   fragments are expected to be clustered around the binding sites,
%   therefore it is less likely to observe this fragments scattered in
%   the genome. In a bad experiment is the opposite case, so the
%   fragments are not clustered so if the enzyme doesn't have a strong
%   preference to how is attracted to the fragments then the remaining
%   ones can be seen as random positions around the genome, this
%   positions are copied several times during the amplification and that
%   may be the reason why we observe something as scattered. Perhaps if
%   the enzyme was left for longer time, then the starting material that
%   later was amplified into this regions would have been digested and
%   some of this regions wouldn't appear (I am thinking about the deeply
%   sequenced TBP experiments.)}

@@

% while there are not established QC pipelines for ChIP-exo; previous
% ChIP-exo analyses used ChIP-Seq samples to compare the resolution
% between experiments \cite{exo1,exoillumina,exo2}.

@@

\subsection*{Publicly available ChIP-exo/nexus and novel \textit{E. coli} ChIP-seq/exo datasets}
We utilized a rich collection of publicly available ChIP-exo/nexus
data from multiple organisms to build and evaluate our quality control
pipeline (Table~\ref{tab:qc}). These include: CTCF factor in human
HeLa cells \cite{exo1}; ER factor in human MCF-7 cells
\cite{exoillumina}; GR factor in IMR90, K562, and U2OS human cells
\cite{starick15}; TBP factor in human K562 cells \cite{venters13}.
ChIP-nexus data included experiments from \cite{chipnexus} profiling
TBP in human K562 cells, MyC and Max in \textit{D. melanogaster} S2
cells, and Twist and Dorsal in \textit{D. melanogaster} embryo.

In order to have a setting where we can compare SE and PE ChIP-seq
with their ChIP-exo counterpart, we profiled $\sig$ under a variety of
conditions in \textit{E. coli} with both ChIP-exo and PE ChIP-seq
(Table~\ref{tab:qc_sig}). Collectively, we generated $\sig$ factor
ChIP-exo, PE and SE ChIP-seq experiments under aerobic ($+O_2$) and
anaerobic ($-O_2$) conditions in glucose minimal media, for simplicity
we named these experiments as E1, P1 and S1 respectively. Similarly,
we generated $\sig$ factor ChIP-exo, PE and SE (generated \textit{in
  silico}) ChIP-seq experiments in \textit{E. coli} under aerobic
($+O_2$) conditions with and without rifampicin treatment.  We refer
to these latter set of experiments as time point 20 minutes and time
point 0 minutes, we also named these experiments E2, P2 and S2 for
ChIP-exo, PE ChIP-seq and SE ChIP-seq respectively.
% \SK{Check for
%   consistency with other parts of the paper and correct, ok to use
%   different naming convention}.
We specifically used these experimental data for comparisons of
ChIP-exo and PE ChIP-seq assays in identifying closely spaced binding
events and their resolution, i.e., physical proximity of the predicted
events to the actual binding sites.  This comparison benefits from
using $\sig$ which is a transcription initiation factor of
housekeeping genes in \textit{E. coli}. Many \textit{E. coli}
promoters contain multiple transcription start sites (TSS). These TSSs
are often closely spaced, i.e., within 10 $\sim$ 150 bps of each
other, and are considered to be multiple ``switches'' that
differentially regulate gene expression under diverse growth
conditions \cite{regulondb}.

@@

 % \SK{Maybe add a column as sample
 %  number to this table to make referencing easy}
\textit{Read distributions within signal and background regions.} We
first compared ChIP-seq and ChIP-exo in terms of data features that
are well studied in ChIP-seq studies.  Our $\sig$ ChIP-seq and
ChIP-exo samples from \textit{E. coli} are especially well suited for
this task since they are all deeply sequenced compared to the genome
size of \textit{E. coli}.  Figure~\ref{fig:1} summarizes this
comparison for one biological replicate of ChIP-exo and ChIP-seq
experiments from the same biological conditions (samples EB1-1 and
P1-1 from Table~\ref{tab:qc_sig}). Comparisons with other paired
\textit{E. coli} ChIP-seq and ChIP-exo samples led to similar
conclusions (Supplementary Figure~\ref{sfig:comp1}).
%compared various factors that could affect binding site
%identification between ChIP-exo and ChIP-Seq data by using the
%ChIP-exo reads from the first biological sample and first replicate
%grown under aerobic condition (first line from Table \ref{tab:qc_sig})
%and a SE ChIP-Seq replicate grown under the same conditions. 
We summarized the extended raw read counts within $150$ bps
non-overlapping intervals, i.e., bins, interrogating the genome.
%In order
%to compare distribution of signal and background between ChIP-exo and
%ChIP-Seq data, we counted the number of extended reads mapping to a
%partition of the genome into non-overlapping bins. 
Figure~\ref{fig:1}B depicts that, as expected, ChIP read counts
from ChIP-exo and ChIP-seq are linearly correlated especially at high
read counts. This indicates that signals for potential binding sites
are well reproducible between ChIP-exo and ChIP-seq data.  In
contrast, there is a clear difference among the two data types for
bins with low read counts, highlighting expected differences in the
background read distributions of the two data types.  \SK{***After
  seeing these types of plots from other samples (e..g., ER in MCF7,
  it is unlikely that what we are seeing from $\sig$ samples regarding
  background reads (the next few sentences) is typical. For CTCF and
  $\sig$, the ChIP-exo samples have a wider dynamic range, but this
  could be due to differences in sequencing depth. *** Specifically,
  background reads from ChIP-seq are almost uniformly distributed over
  background (non-enriched) regions. In comparison, background regions
  in ChIP-exo show larger variation but have overall lower read
  counts. Furthermore, there are considerably large numbers of bins
  with zero read counts in ChIP-exo and non-zero read counts in
  ChIP-seq.
%and the
%ChIP tag counts in there regions were significantly larger than
%zero. In contrast, in ChIP-exo data, there was larger variation in
%ChIP tag counts among background regions and ChIP tag counts were much
%lower in these regions compared to ChIP-Seq data. There were also
%large proportion of regions without any read in ChIP-exo data. 
  Overall, these observations indicate that a much smaller portion of
  the genome is expected to be background in ChIP-exo compared to
  ChIP-seq and methods that specifically model the background read
  distribution might benefit from acknowledging this.} \RW{Figure 2B
  from the chip-nexus paper makes a similar comparison but specific to
  peaks, i.e. our argument about both ChIP-seq and ChIP-exo is
  correlated in high signal region still holds.}

\textit{Peak-pair assumption.} We next evaluated the peak-pair
assumption, i.e., a cluster of reads in the forward strand is usually
paired with a cluster of reads in the reverse strand that is located
on the right-hand-side of the binding site, that is commonly utilized
in designing statistical analysis methods for ChIP-exo data
\cite{mace, cexor, peakzilla}.  We considered the set of peaks
identified in both the ChIP-seq and ChIP-exo samples as high quality
peaks (Materials and Methods) and calculated the proportion of forward
strand reads in these regions (Figure~\ref{fig:1}C). This plot
reveals a higher level of strand imbalance for ChIP-exo compared to
ChIP-seq.
% In order to evaluate this assumption, we reviewed the proportion of
% reads in the forward strand in high quality ChIP-exo peaks such as
% at least one binding site is predicted in both ChIP-exo and ChIP-Seq
% data.  We found that strands of reads were much less balanced in
% ChIP-exo data than in ChIP-Seq data in these regions with potential
% binding sites (Fig. \ref{fig:comp}B) and this indicates that the
% peak pair assumption might not hold in ChIP-exo data.
Potential reasons for this observation include ligation efficiency,
efficiency of $\lambda$-exo digestion, and single-stranded protein-DNA
interactions. Overall, such an imbalance is prevalent in 85\% of the
ChIP-exo samples used in this paper \RW{(Supplement
  Figures~\ref{sfig:comp1} to \ref{sfig:comp3})}. \RW{Note: The only ChIP-exo
  experiment were it seems that ChIP-exo is not more imbalanced than
  ChIP-seq is Carroll's ER dataset.}
%the enzyme digestion or the strength at which the protein binds to the
%DNA: Although it is expected for the exonuclease enzyme to digest the
%ChIP fragments starting by their $5\prime$ ends and stopping when
%finding a binding event, it may not be able to reach and digest every
%fragment in the ChIP sample. On the other hand, if the protein is not
%bound to both DNA strand, the enzyme may completely digest the
%fragment. Finally both effect are increased by the PCR amplification.

\textit{Mappability and GC-content bias.} We next evaluated ChIP-exo
data of CTCF in HeLa cells \cite{exo1} to investigate biases inherent
to next generation sequencing experiments with eukaryotic genomes.
Figures \ref{fig:1}E and \ref{fig:1}F display the bin-level
average read counts against mappability and GC-content. Each data
point is obtained by averaging the read counts across bins with the
same mappability of GC-content.  These biases, increasing linear trend
with mappability and non-linear trend with GC-content, are similar to
those observed in ChIP-seq datasets \cite{benjamini2011,
  mosaics,quest}.
%In Figure \ref{fig:comp}C it is shown that the ChIP-exo tag
%counts linearly increases with the mappability score and in Figure
%\ref{fig:comp}D it is shown that for GC - content below 0.6, the mean
%ChIP tag count increases and for GC - content greater than 0.6 it
%shows a decreasing trend. Benjamini and Speed,
%2011\nocite{benjamini2011} and Kuan et al., 2011\nocite{mosaics}
%studied the presence of the mappability and GC - content biases in
%ChIP-Seq's background. It is not surprising to see these biases also
%present in ChIP-exo data, since ChIP-Seq and ChIP-exo signal seems to
%be linearly correlated for enriched regions (Figure
%\ref{fig:comp}A). Rozowsky et al., 2009\nocite{peakseq} and Valouev et
%al., 2008\nocite{quest} provide in depth analysis of the mappability
%and GC - content biases for ChIP-Seq respectively. 
This observation  indicates that analysis of  ChIP-exo data
should benefit from methods that take into account  apparent
sequencing biases such as mappability and GC content, mostly when an
input  control sample is not available.

@@

The Strand Cross-Correlation (SCC), introduced by \cite{strandcc}, is the most commonly used quality metric in assessing 
ChIP-seq enrichment quality. It aims to quantify  how well the reads mapped to each
strand are clustered around the locations of the protein-DNA interaction sites by calculating the Pearson correlation among forward and backward strands reads by shifting them across a range that covers both the read length of the experiment and the expected average fragment length.
Typical SCC profiles exhibit two local maxima: at the average fragment length and the read length. In high quality experiments with clear ChIP enrichment, the average fragment length maximum coincides with the global maximum.
%, and usually it is expected to observed two local
%maxima, one when the profiles are shifted by the average read length
%and another when the profiles are shifted by the unobserved fragment
%length. 
%In a high quality ChIP-Seq dataset the last one is also the
%SCC global maxima. 
In an idealized ChIP-exo experiment where the DNA fragments are
digested to the boundaries of the protein-DNA interaction sites, we
would expect the SCC profile to maximize at the motif length
indicating clustering of the forward and reverse strand reads around
the binding site. Figure~\ref{fig:1}D displays the SCC curves for the
CTCF HeLa samples where the ChIP-exo curve shows local maxima at the
motif and read lengths, while the SE ChIP-seq curves have a local
maxima at the read length and a global maxima at the average fragment
length. SCC profiles for other samples are available in Supplementary
Figures \ref{sfig:scc1} to \ref{sfig:scc8}. The read length and motif
length maxima are often in close proximity of each other; as a result,
this renders QC metrics such as the Normalized Strand
Cross-Correlation (NSC) or the Relative Strand Cross-Correlation (RSC)
harder to interpret; however, the profile itself seems informative
about the enrichment signal in ChIP-exo/nexus experiments \SK{this
  seems true at least based on my initial look at the plots, but we
  should be cautious in reporting this}. \RW{i agree, and additionally
  Carroll et al, 2014 kinda agrees with us that the SCC-related
  measure may not be useful since the fragment length coincides with
  read length}

 % %http://journal.frontiersin.org/article/10.3389/fgene.2014.00075/full#h5
 %  the paragraph is a little bit above conclusions

@@

We specifically considered recently developed Peakzilla
\cite{peakzilla}, MACE \cite{mace}, GEM \cite{gem}, and dPeak
\cite{dpeak} in our evaluations with high quality \textit{E. coli}
samples (samples E1-1 and E1-2). Among these methods, Peakzilla and
MACE are specifically developed for ChIP-exo analysis, whereas both
dPeak and GEM can identify high resolution binding events from ChIP
experiments, rendering them suitable for ChIP-exo analysis as well.
%After we determined which biases could affect a ChIP-exo experiment's
%quality we decided to proceed and examine binding sites in high
%resolution on high quality ChIP-exo experiments. We examined first
%both aerobic replicates from the \emph{E.Coli} first biosample.

Figures~\ref{fig:6}A and \ref{fig:6}B compare binding events
identified with all the four methods in terms of resolution, where the
resolution is defined as the distance from a RegulonDB
\cite{regulondb} annotation to its closest prediction. The resolutions
of all the methods are comparable for the first replicate
(Figure~\ref{fig:6}A) and dPeak, on average, has slightly better
resolution than the rest in the second replicate
(Figure~\ref{fig:6}B).
